% 中文摘要
\chapter[摘要]{摘\quad 要} 
\setlength{\parindent}{2em}

随着元宇宙、虚拟现实及数字娱乐产业的飞速发展，构建高保真、可驱动且具备结构化语义的数字化身已成为计算机视觉与图形学领域的研究热点。然而，穿衣数字化身的重建与生成仍面临两大瓶颈：其一，现有基于三维高斯泼溅的人体重建方法中，高斯基元的旋转与尺度属性通过自由优化获得，与底层网格拓扑解耦，导致在分布外姿态驱动时出现表面破碎与伪影；其二，现有参数化服装生成方法缺乏对人体几何形态的显式感知，生成的服装版型难以适配特定体型，且严重依赖大规模训练数据。针对上述问题，本文以"显式几何先验"为核心线索，开展以下研究工作：

针对 3D 高斯泼溅人体重建在新姿态下的几何失真问题，本文提出了一种拓扑感知的三维高斯人体重建方法。该方法利用可变形网格的拓扑结构，设计了面片高斯与顶点高斯两类互补基元，分别以三角面片内切圆和网格顶点为锚点，所有几何属性均从网格拓扑解析推导，从根本上消除了高斯与表面的脱耦问题；同时引入等边几何正则化以维护网格质量。在 PeopleSnapshot 与 X-Avatar 数据集上的实验表明，相较于最优基线，该方法将法向误差降低了 13.8\%--17.9\%，倒角距离降低了 15.9\%，模型体积缩减 44.5\%。

针对现有服装参数生成方法忽视人体尺寸约束且依赖大规模训练的缺陷，本文提出了一种基于人体几何先验的服装参数生成方法。该方法首先将量体工具适配至 SMPL 拓扑以从单张图像提取人体测量值作为几何先验，继而通过信息分流策略将参数格式压缩至紧凑表示以降低超过 90\% 的 Token 消耗，最终通过几何锚点归一化与分步推理工作流引导 GPT-4o 以零样本方式完成多模态参数推理。在 CLoSe 与 4D-DRESS 数据集上的实验表明，该方法分别取得 1.621 cm 和 1.834 cm 的倒角距离，较最优基线降低 19.5\% 和 6.2\%，推理速度提升约 4 倍，且无需训练数据，结构有效率达 100\%。

综上所述，本文通过引入不同层级的几何先验，分别解决了数字化身在动态外观渲染中的几何失真问题和参数化服装生成中的体型失配问题，为构建高质量、可交互的穿衣数字化身提供了有效的技术支撑。

\par \vspace{1em} % 空一行
\noindent \textbf{关键词：}数字化身重建；3D 高斯泼溅；参数化服装生成；几何先验


% 英文摘要
\chapter[Abstract]{Abstract}
% \setlength{\parindent}{0em} % 英文摘要通常首行不缩进，段落间空行；若学校有格式要求可改为 2em

With the rapid development of the Metaverse, virtual reality, and digital entertainment industries, constructing high-fidelity, drivable, and semantically structured digital avatars has become a research hotspot in computer vision and graphics. However, the reconstruction and generation of clothed digital avatars still face two key bottlenecks: first, in existing 3D Gaussian Splatting-based human reconstruction methods, the rotation and scale attributes of Gaussian primitives are freely optimized and decoupled from the underlying mesh topology, causing surface collapse and artifacts under out-of-distribution poses; second, existing parametric garment generation methods lack explicit perception of human body geometry, resulting in ill-fitting garment patterns and heavy reliance on large-scale training data. To address these issues, this thesis takes ``explicit geometric priors'' as the core thread and conducts research from two dimensions: appearance reconstruction and structure generation. The main contributions are as follows:

To address geometric distortion in 3D Gaussian Splatting-based human reconstruction under novel poses, this thesis proposes a topology-aware 3D Gaussian human reconstruction method. The method designs two complementary Gaussian primitive types---face Gaussians and vertex Gaussians---anchored at triangle incircles and mesh vertices respectively, with all geometric attributes analytically derived from mesh topology, fundamentally eliminating the decoupling between Gaussians and the surface; an equilateral geometric regularization term is further introduced to maintain mesh quality. Experiments on PeopleSnapshot and X-Avatar show that, compared to the best baseline, the method reduces normal error by 13.8\%--17.9\% and Chamfer distance by 15.9\%, with 44.5\% smaller model size.

To address the deficiencies of existing garment parameter generation methods that neglect body size constraints and rely on large-scale training, this thesis proposes a body-geometry-prior-guided garment parameter generation method. The method first adapts a body measurement tool to SMPL topology to extract anatomical measurements as geometric priors from a single image, then compresses the parameter format into a compact representation via an information routing strategy to reduce token consumption by over 90\%, and finally guides GPT-4o to perform zero-shot multimodal parameter reasoning through geometric anchor normalization and a step-by-step workflow. Experiments on CLoSe and 4D-DRESS show that the method achieves Chamfer distances of 1.621\,cm and 1.834\,cm respectively, outperforming the best baselines by 19.5\% and 6.2\%, with approximately 4$\times$ faster inference, no training data required, and a 100\% structural validity rate.

In summary, by introducing geometric priors at different levels, this thesis addresses both the geometric distortion problem in dynamic appearance rendering and the body mismatch problem in parametric garment generation for digital avatars, providing effective technical support for building high-quality and interactive clothed digital avatars.

\par \vspace{1em}
\noindent \textbf{Keywords:} Digital Avatar Reconstruction; 3D Gaussian Splatting; Parametric Garment Generation; Geometric Priors

\setlength{\parindent}{2em} 